---
title: "body_macroregion_plotting"
author: "Paul Gaudin"
date: '2024-04-02'
output: html_document
---

## Load Necessary Libraries

```{r}

# to install updated CHOIRBM, navigate into CHOIRBM directory ( setwd('CHOIRBM') ) and then run devtools::install()
library(CHOIRBM)
# - use to check package version - should print ‘0.0.2.9001’ if updated CHOIRBM packaged loaded
packageVersion("CHOIRBM") 


# Load other necessary packages
# install.packages(c("readxl", "dplyr", "tidyr", "stringr", "gt", "webshot2")) # webshot2 needed for saving gt tables
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(gt)
library(ggplot2)

```

## Create Necessary Lists


```{r}

## color palette that will be used in macroregion plotting. 1st will show as region with highest percentage, last with lowest
new_color_palette <- c(
  "#F8BBD0", # Pastel Red
  "#FFCCBC", # Pastel Orange
  "#FFF9C4", # Pastel Yellow
  "#C8E6C9", # Pastel Light Green
  "#B2DFDB", # Pastel Green-Blue (Aqua)
  "#B3E5FC", # Pastel Blue
  "#BBDEFB", # Lighter Blue
  "#D1C4E9", # Pastel Purple
  "#F3E5F5", # Lighter Purple
  "#F5F5F5", # Very Light Grey
  "#FFFFFF"  # White
)

## map to assign microregions to macroregions
body_map <- list(
  "head_front" = c(101, 102, 103, 104),
  "head_back" = c(201, 202, 203, 204),
  "neck" = c(105, 106, 205, 206),
  "right_shoulder_arm" = c(107, 111, 115, 117, 119, 125, 210, 214, 216, 220, 226, 230),
  "left_should_arm" = c(110, 114, 116, 118, 124, 128, 207, 211, 215, 217, 221, 227),
  "upper_back" = c(208, 209, 212, 213),
  "lower_back" = c(218, 219),
  "chest_abdomen" = c(108, 109, 112, 113),
  "pelvic" = c(121, 122),
  "right_hip_buttock_leg" = c(120, 126, 129, 131, 133, 135, 224, 225, 229, 232, 234, 236, 238),
  "left_hip_buttock_leg" = c(123, 127, 130, 132, 134, 136, 222, 223, 228, 231, 233, 235, 237)
)

## map for naming conventions for macroregions
body_map_pain_macroregion_format_list <- list(
  "head_front" = "Head Front",
  "head_back" = "Head Back",
  "neck" = "Neck",
  "right_shoulder_arm" = "Right Shoulder/Arm",
  "left_should_arm" = "Left Shoulder/Arm",
  "upper_back" = "Upper Back",
  "lower_back" = "Lower Back",
  "chest_abdomen" = "Chest/Abdomen",
  "pelvic" = "Pelvic",
  "right_hip_buttock_leg" = "Right Hip/Buttock/Leg",
  "left_hip_buttock_leg" = "Left Hip/Buttock/Leg"
)



```

## Import Data from Excel file

```{r}
# Import the Excel file
data <- read_excel("lupus body map counts by gender.xlsx")

```

## Create Variables and Define Necessary Functions for Male and Female Tables 

```{r}

## list of all numbers - necessary in case any are missing from male/female data 
all_body_map_numbers <- unique(unlist(body_map))


calculateRegionOccurrences <- function(df, body_map, gender) {
  # This function is used to determine total occurrences of any number within macroregion for each patient, and the percent occurrences
  #
  # Parameters:
  #   df: the data.frame from the excel file 
  #   body_map: the body_map list
  #   gender: string "Female" or "Male" 
  #
  # Returns:
  #   2 named lists, total_occurrences and percent_occurrences
  #

  # Filter DataFrame for gender
  df_gender <- df[df$patient_gender == gender, ]
  
  # List to hold the occurrence counts for each region, starting with 0
  region_occurrences <- sapply(names(body_map), function(x) 0)
  
  # Iterate over each row in df_gender (filtered data by gender) and add occurrence if microregion within macroregion present
  for (i in 1:nrow(df_gender)) {
    # Split the "bodymap_regions_csv" string into a vector of numbers
    regions_vector <- as.numeric(unlist(strsplit(as.character(df_gender$bodymap_regions_csv[i]), ",")))
    # Check each body map region for occurrence
    for (region_name in names(body_map)) {
      region_ids <- body_map[[region_name]]
      # Add if any region ID is present in the regions_vector
      if (any(region_ids %in% regions_vector)) {
        region_occurrences[region_name] <- region_occurrences[region_name] + 1
      }
    }
  }
  
  # Calculate the perfect occurrences for each region
  percent_occurrences <- sapply(region_occurrences, function(x) round((x / nrow(df_gender) * 100), 2))
  
  return(list(
    total_occurrences = region_occurrences,
    percent_occurrences = percent_occurrences
  ))
}

## function 
getGenderTables <- function(data, gender, body_map, body_map_pain_macroregion_format_list, new_color_palette) {
  # This function is used to generate the formatted tables with gt package, as well the gender_region_occurrences named list that will be used in the body image plotting
  #
  # Parameters:
  #   data: the excel data.frame
  #   gender: "Male" or "Female" string 
  #   body_map: the body_map named list 
  #   body_map_pain_macroregion_format_list: the named list that converts the names in body_map to their display names 
  #   new_color_palette: the color palette used for the tables and body images generated
  #
  # Returns:
  #   Named list, with Table being the gt generated table, and Occurrences being the named list with the total and percent occurences as sub named lists
  #
  
  ## calculate and assign the total and percent occurrences
  gender_region_occurrences <- calculateRegionOccurrences(data, body_map, gender)
  
  ## assign the properly formatted names for the table 
  body_map_regions <- sapply(names(body_map_pain_macroregion_format_list), function(key) body_map_pain_macroregion_format_list[[key]])
  ## assign the percent occurences
  overall_percentages <- gender_region_occurrences$percent_occurrences[names(body_map_pain_macroregion_format_list)]

  # create data.frame with necessary the :Body Map Pain Region" and "Overall" columns
  pain_regions_df <- data.frame(
    "Body Map Pain Regions" = body_map_regions,
    "Overall" = overall_percentages,
    stringsAsFactors = FALSE
  )
  
  # assign Ranks and Colors for each of the rows, so colors will reflect the percents
  pain_regions_df$Rank <- rank(-pain_regions_df$Overall, ties.method = "first")
  pain_regions_df$Color <- new_color_palette[pain_regions_df$Rank]

  # calculate total indivuals and create label for "Overall" column 
  num_individuals <- nrow(data[data$patient_gender == gender, ])
  overall_label <- paste("Overall (n=", num_individuals, ")", " ", gender, sep = "")

  # create gt_table 
  gt_table <- gt(pain_regions_df) %>%
    cols_label(
      Body.Map.Pain.Regions = "Body Map Pain Regions",
      Overall = overall_label 
    ) %>%
    tab_options(column_labels.font.weight = "bold")



  # Format the "Overall" column to include a percentage sign for the percents
  gt_table <- gt_table %>%
    fmt_number(
      columns = c(Overall),
      decimals = 2, # Assuming you want two decimal places; adjust as needed
      suffixing = TRUE,
      pattern = "{x}%"
    )

  # Add borders to the gt_table 
  gt_table <- gt_table %>%
    # Apply borders to the body of the table
    tab_style(
      style = cell_borders(
        sides = "all", # Apply borders to all sides of each cell in the body
        color = "gray",
        weight = px(1)
      ),
      locations = cells_body(columns = everything(), rows = TRUE)
    ) %>%
    # Apply borders to the column labels
    tab_style(
      style = cell_borders(
        sides = "all", # Apply borders to all sides of the column label cells
        color = "gray",
        weight = px(1)
      ),
      locations = cells_column_labels(columns = everything())
    ) %>%
    # Apply borders to the table title
    tab_style(
      style = cell_borders(
        sides = "all", # Apply borders around the table title
        color = "gray",
        weight = px(1)
      ),
      locations = cells_title(groups = "title")
    ) 
  
  # Apply colors to the gt_table row-wise based on 'Rank'
  for (rank in unique(pain_regions_df$Rank)) {
    color <- pain_regions_df$Color[pain_regions_df$Rank == rank][1]
    gt_table <- gt_table %>%
      tab_style(
        style = cell_fill(color = color),
        locations = cells_body(
          columns = everything(),
          rows = pain_regions_df$Rank == rank
        )
      )
  }

  # Hide the 'Rank' and 'Color' columns
  gt_table <- gt_table %>%
    cols_hide(columns = c(Rank, Color))
  

  # return the gt_table and gender_region_occurrences
  return(list(Table = gt_table, Occurrences = gender_region_occurrences))

}
```

## Assign and display the results for the Female pain macroregions

```{r}
female_result <- getGenderTables(data, "Female", body_map, body_map_pain_macroregion_format_list, new_color_palette)
female_table <- female_result$Table
female_region_occurrences <- female_result$Occurrences ## track female_region_occurrences for use in body image plot
female_table
```

## Assign and display the results for the Male pain macroregions

```{r}
male_result <- getGenderTables(data, "Male", body_map, body_map_pain_macroregion_format_list, new_color_palette)
male_table <- male_result$Table
male_region_occurrences <- male_result$Occurrences
male_table
```

## Functions for formatting total counts from the excel data to be used in the CHOIRBM package

```{r}
getGenderDataCounts <- function(data, gender, all_body_map_numbers){
  # This function is used to generate a data.frame that includes the microregion id, value, and group columns, necessary for plotting with CHOIRBM
  #
  # Parameters:
  #   data: the excel data.frame
  #   gender: "Male" or "Female" string 
  #   all_body_map_numbers: flattened list containing all microregion numbers
  #
  # Returns:
  #   data.frame formatted for input with microregion id, value, and group
  
  # Create data.frame, filtered for gender
  gender_data <- data %>% 
    filter(patient_gender == gender)
  
  # Split the "bodymap_regions_csv" column into individual numbers, then count each occurrence
  number_counts <- gender_data %>%
    separate_rows(bodymap_regions_csv, sep = ",") %>%
    group_by(bodymap_regions_csv) %>%
    summarize(occurrence = n()) %>%
    ungroup()

  # Create a new dataframe with value as the the occurence count
  gender_data_totals <- number_counts %>%
    mutate(id = as.integer(bodymap_regions_csv),
           value = occurrence) %>%
    select(id, value)

  # Check which numbers are not in gender_data_totals$id and add them with value = 0
  missing_numbers <- all_body_map_numbers[!all_body_map_numbers %in% gender_data_totals$id]
  missing_data <- data.frame(id = missing_numbers, value = rep(0, length(missing_numbers)))

  # Combine the original dataframe with the missing data
  final_gender_data_counts <- rbind(gender_data_totals, missing_data)
  
  # Ensure the dataframe is ordered by 'id' after adding missing numbers
  final_gender_data_counts <- final_gender_data_counts[order(final_gender_data_counts$id),]
  
  # Add 'group' column based on the first digit of 'id' - 1 is Front, 2 is Back 
  final_gender_data_counts <- final_gender_data_counts %>%
    mutate(group = factor(ifelse(substr(as.character(id), 1, 1) == "1", "Front", "Back")))
  
  # Make sure the groups are defined as Levels - necessary for correct plotting with CHOIRBM 
  final_gender_data_counts$group <- factor(final_gender_data_counts$group, levels = c("Front", "Back"))
  
  # Assign id as character value
  final_gender_data_counts <- final_gender_data_counts %>%
    mutate(id = as.character(id))
  
  # Remove any NA values
  final_gender_data_counts <- final_gender_data_counts %>%
    filter(!is.na(id))
  
  # Assign as regular data.frame, not tidy format
  final_gender_data_counts <- as.data.frame(final_gender_data_counts)
                                   
  return(final_gender_data_counts)             
}

```

## Generate Female Data Counts and Plot Body Macroregion Map 

```{r}
female_data_counts <- getGenderDataCounts(data, "Female", all_body_map_numbers)
# plot_female_macroregions_with_counts is a new function in the CHOIRBM package. find under CHOIRBM/R/plot_female_macroregions_with_counts.R
female_plot <- plot_female_macroregions_with_counts(female_data_counts, "value", female_region_occurrences$percent_occurrences, new_color_palette, body_map)
female_plot
```

## Generate Male Data Counts and Plot Body Macroregion Map 

```{r}
male_data_counts <- getGenderDataCounts(data, "Male", all_body_map_numbers)
# plot_male_macroregions_with_counts is a new function in the CHOIRBM package. find under CHOIRBM/R/plot_male_macroregions_with_counts.R
male_plot <- plot_male_macroregions_with_counts(male_data_counts, "value", male_region_occurrences$percent_occurrences, new_color_palette, body_map)
male_plot
```

## Create data.frame with Total Occurrences, and Create Total Occurrence table 

```{r}

## assign totals 
n_total <- nrow(data)
n_female <- sum(data$patient_gender == "Female")
n_male <- sum(data$patient_gender == "Male")

# Map the region names to the full labels
full_labels <- sapply(names(body_map_pain_macroregion_format_list), 
                      function(x) paste(body_map_pain_macroregion_format_list[[x]], "Region"))

# Calculate Overall percentages
overall_percentages <- (female_region_occurrences$total_occurrences + male_region_occurrences$total_occurrences) / n_total * 100

# Create a combined dataframe with occurrences and percentages
combined_data <- data.frame(
  BodymapRegions = names(female_region_occurrences$total_occurrences),
  Overall = paste(female_region_occurrences$total_occurrences + male_region_occurrences$total_occurrences, " (", sprintf("%.1f", overall_percentages), ")", sep = ""),
  Female = paste(female_region_occurrences$total_occurrences, " (", sprintf("%.1f", female_region_occurrences$percent_occurrences), ")", sep = ""),
  Male = paste(male_region_occurrences$total_occurrences, " (", sprintf("%.1f", male_region_occurrences$percent_occurrences), ")", sep = "")
)

# Map the region names to the full labels with " Region" appended
combined_data$BodymapRegions <- sapply(combined_data$BodymapRegions, function(x) {
  paste(body_map_pain_macroregion_format_list[[x]], "Region")
})

# Create the GT table
overall_table <- gt(combined_data) %>%
 cols_label(
    BodymapRegions = "Bodymap Regions with reported pain",
    Overall = paste("Overall (n=", n_total, ")", sep = ""),
    Female = paste("Female (n=", n_female, ")", sep = ""),
    Male = paste("Male (n=", n_male, ")", sep = "")
  ) %>%
  tab_options(column_labels.font.weight = "bold") %>%
  fmt_passthrough(columns = c(Overall, Female, Male))  # Use fmt_passthrough to keep the formatting as is

## Add borders to the table
overall_table <- overall_table %>%
  # Apply borders to the body of the table
  tab_style(
    style = cell_borders(
      sides = "all", # Apply borders to all sides of each cell in the body
      color = "gray",
      weight = px(1)
    ),
    locations = cells_body(columns = everything(), rows = TRUE)
  ) %>%
  # Apply borders to the column labels
  tab_style(
    style = cell_borders(
      sides = "all", # Apply borders to all sides of the column label cells
      color = "gray",
      weight = px(1)
    ),
    locations = cells_column_labels(columns = everything())
  ) %>%
  # Apply borders to the table title
  tab_style(
    style = cell_borders(
      sides = "all", # Apply borders around the table title
      color = "gray",
      weight = px(1)
    ),
    locations = cells_title(groups = "title")
  )

# Print the GT table
overall_table

```

## Save Table and Body Macroregion Images in outputs folder

```{r}
ggsave("outputs/female_plot.png", plot = female_plot, width = 10, height = 6, dpi = 300) ## adjust dpi (dots per inch) to change resolution (higher means higher res)
gtsave(female_table, filename = "outputs/female_table.png")
ggsave("outputs/male_plot.png", plot = male_plot, width = 10, height = 6, dpi = 300)
gtsave(male_table, filename = "outputs/male_table.png")
gtsave(overall_table, filename = "outputs/overall_table.png")
```
